The Dictionary for characters of smooth group schemes

### 2016-03-28 followup.tex
Clarified the proof of Proposition {prop:Gder_triv}.

The first aim of this paper is to repair this defect in the function-sheaf dictionary by describing a full subcategory $\CCS(G)$ of $\CS(G)$ so that $\TrFrob{G}$ restricts to an isomorphism $\CCSiso{G} \to G(k)^*$.
We refer to the resulting objects as \emph{commutative character sheaves}, since the passage from $\CS(G)$ to $\CCS(G)$ involves a condition that exchanges the inputs to the multiplication morphism on $G$ (see Definition \ref{def:CCScom}).  
When $G$ is connected, all character sheaves on $G$ are commutative.
This category clarifies several questions about character sheaves on $G$: invisible character sheaves \cite{cunningham-roe:13a}*{Def. 2.8} are precisely those that are not commutative, and  $\TrFrob{G}^{-1} : G(k)^* \to \CCSiso{G}$ provides a canonical splitting of $\TrFrob{G} : \CSiso{G} \to G(k)^*$ \cite{cunningham-roe:13a}*{Rem. 3.7}.

Next, we broaden our scope further to encompass smooth group schemes $G$ that are not necessarily commutative.
In this paper, we do not consider Lusztig's geometrization of arbitrary characters of connected, reductive groups $G(k)$
using character sheaves \cite{lusztig:85a}*{Def.~2.10}, but focus on the case of $1$-dimensional characters
$G(k)^*$ assuming only that $G$ is smooth, allowing for the possibility that $G$ is not connected and not reductive.
The category $\CS(G)$ has a straightforward generalization to this case, but again
there are more linear character sheaves than there are characters, as pointed out by Kamgarpour \cite{kamgarpour:09a}*{(1.1)}.
We define a category $\CCS(G)$ and a forgetful functor to $\CS(G)$ so that $\TrFrob{G} : \CCSiso{G} \to G\ab(k)^*$
is an isomorphism.  
Since $G\ab(k)^*$ surjects onto $G(k)^*$, it follows that for each character $\chi \in G(k)^*$ there is a commutative
character sheaf $\cs{L}$ on $G$ with $\TrFrob{G}(\cs{L}) = \chi$. Moreover, we find that pullback along the quotient $q : G \to G\ab$
defines an equivalence of categories $\CCS(G\ab) \to \CCS(G)$.  Note that the functor $\CCS(G) \to \CS(G)$ is not
essentially surjective, missing the kinds of linear character sheaves highlighted by Kamgarpour.

@@

If $\chi \in G(k)^*$ is a character, it must vanish on the derived subgroup $G\der(k)$.
Kamgarpour gives an example \cite{kamgarpour:09a}*{(1.1)} of a character sheaf
that does not vanish on $G\der$, defined by the extension
\[
1 \to \mu_n \to \SL_n \to \PGL_n \to 1.
\]
In order to obtain a relationship between character sheaves on $G$ and characters of $G(k)$,
he opts to give a different definition of commutator.  Since we already need to adapt our
notion of character sheaf, even in the commutative case, we instead add restrictions
to the definition of commutative character sheaf and leave the definition of $G\der$ unchanged.

\subsection{Definition}\label{ssec:noncomdef}

In order to get character sheaves that correspond to characters in $G(k)^*$, we must discard those
character sheaves whose restriction to the derived subgroup is nontrivial.
The following proposition will be proven in Section \ref{ssec:proof1}.
\begin{proposition} \label{prop:Gder_triv}
Suppose $G$ is a smooth group scheme and $\cs{L} \in \CS(G)$ is a linear character sheaf on $G$.
Then the restriction of $\cs{L}$ to $G\der$ is trivial if and only if $\cs{L} \cong q^*(\cs{L}\ab)$ for some
character sheaf $\cs{L}\ab$ on $G\ab$.
\end{proposition}

@@

\begin{corollary}\label{cor:Gab}
%Let $G$ be a smooth group scheme over a finite field $\Fq$. Let $\CCS(G)$ be the category of {\it commutative characters sheaves} on $G$; see Definition~\ref{def:CCS}.
Category $\CCS(G)$ is monoidal and $\CCSiso{G} \iso \Hom(G\ab(\Fq),\EEx)$, canonically.
Every map in $\CCS(G)$ is either trivial or an isomorphism, and the automorphism group of any object in $\CCS(G)$ is $\Hom(\pi_0(G\ab)_{\Frob{}},\EEx)$, canonically.
\end{corollary}

@@

%\begin{proof}[Proof of Proposition~\ref{prop:Gder_triv}] 
By \cite{demazure:SGA3-VIA}*{Thm. 3.2}, the abelianization $G\ab\ceq G/G\der$ exists in schemes over $\Fq$ and the quotient $q: G \to G\ab$ is a $G\der$-torsor in the fppf topology.
%
To simplify notation below, set $H=G\der$
and let $j : H\hookrightarrow G$ be the inclusion. 
With reference to \eqref{qH} and Section~\ref{ssec:closed}, consider the following diagram.
%\[
%\begin{tikzcd}
%{} & G\times G \arrow{d}{c} &&&\\
%1 \arrow{r} & H \arrow{r}{j} & G \arrow{r}{q} & G\ab \arrow{r} & 1
%\end{tikzcd}
%\]
%gives
\[
\begin{tikzcd}
\CS(H) & \arrow{l}[swap]{j^*} \CS(G) &\arrow{l}[swap]{q^*} 
\CS(G\ab) \arrow{dl}{\text{equiv}}[swap]{q_{H}^*} \\
 \CS'(G)  \arrow{r}{\beta \mapsto \nu}[swap]{\text{equiv}} &  \CS_{H}(G) \arrow{u}{\text{forget}}  & 
\end{tikzcd}
\]
%
Here we use the notation $\CS'(G)$ from Section~\ref{ssec:noncomdef} and the observation that $\CS'(G)$ is precisely the category $\CS^{H}(G)$ from Section~\ref{ssec:closed} for $H=G\der$.
%
Since the quotient $q : G \to G\ab$ is an $H$-torsor in the fppf topology \cite{demazure:SGA3-VIA}*{Thm. 3.2}, 
it is also an $H$-torsor in the fpqc topology.
It now follows from Lemma~\ref{lem:torsor} that the comparison functor $\CS(G\ab)\to \CS_{H}(G)$ is an equivalence.
%\todo{Did I compare these topologies correctly? fpqc is finer than fppf}
%
On the other hand, by Lemma~\ref{lem:beta}, $\CS'(G) \to \CS_{H}(G)$, defined by $(\cs{L},\mu,\beta)\mapsto (\cs{L},\mu, \nu)$, is an equivalence.
This completes the proof of Proposition~\ref{prop:Gder_triv}.
%\end{proof}

>>>>

The first aim of this paper is to repair this defect in the function-sheaf dictionary for characters of smooth commutative group schemes $G$ by describing a full subcategory $\CCS(G)$ of $\CS(G)$ so that $\TrFrob{G}$ restricts to an isomorphism $\CCSiso{G} \to G(k)^*$.
We refer to the resulting objects as \emph{commutative character sheaves}, since the passage from $\CS(G)$ to $\CCS(G)$ involves a condition that exchanges the inputs to the multiplication morphism on $G$ (see Definition \ref{def:CCScom}).  
When $G$ is connected, all character sheaves on $G$ are commutative.
This category clarifies several questions about character sheaves on $G$: invisible character sheaves \cite{cunningham-roe:13a}*{Def. 2.8} are precisely those that are not commutative, and $\TrFrob{G}^{-1} : G(k)^* \to \CCSiso{G}$ provides a canonical splitting of $\TrFrob{G} : \CSiso{G} \to G(k)^*$ \cite{cunningham-roe:13a}*{Rem. 3.7}.

Next, we broaden our scope further to encompass smooth group schemes $G$ over $\Fq$ that are not necessarily commutative.
In this paper, we do not consider Lusztig's geometrization of characters of representations of connected, reductive groups $G(k)$
using character sheaves \cite{lusztig:85a}*{Def.~2.10}, but focus on the case of characters
$G(k)^*$ (that is, group homomorphisms from $G(k)$ to $\EEx$), assuming only that $G$ is smooth, allowing for the possibility that $G$ is not connected and not reductive.
The category $\CS(G)$ has a straightforward generalization to this case, but again
there are more character sheaves than there are characters, as pointed out by Kamgarpour \cite{kamgarpour:09a}*{(1.1)}.
We define a category $\CCS(G)$ and a forgetful functor to $\CS(G)$ so that $\TrFrob{G} : \CCSiso{G} \to G\ab(k)^*$
is an isomorphism.  
Since $G\ab(k)^*$ surjects onto $G(k)^*$, it follows that for each character $\chi \in G(k)^*$ there is a commutative
character sheaf $\cs{L}$ on $G$ with $\TrFrob{G}(\cs{L}) = \chi$. Moreover, we find that pullback along the quotient $q : G \to G\ab$
defines an equivalence of categories $\CCS(G\ab) \to \CCS(G)$.  Note that the functor $\CCS(G) \to \CS(G)$ is not
essentially surjective, missing the kinds of linear character sheaves highlighted by Kamgarpour.

@@

If $\chi \in G(k)^*$ is a character, it must vanish on the derived subgroup $G\der(k)$.
Kamgarpour gives an example \cite{kamgarpour:09a}*{(1.1)} of a character sheaf
that does not vanish on $G\der$, defined by the extension
\[
1 \to \mu_n \to \SL_n \to \PGL_n \to 1.
\]
In order to obtain a relationship between character sheaves on $G$ and characters of $G(k)$,
he opts to give a different definition of commutator and, in doing so, introduces a `stacky abelianization' of $G$ in order to geometrize characters of $G(\Fq)$.
 Since we has already seen the need to adapt the
notion of character sheaf, even in the commutative case, we instead add restrictions
to the definition of commutative character sheaf and leave the definition of $G\der$ unchanged, allowing us to use the schematic abelianization of $G$ in the geometrization of characters of $G(\Fq)$; see Theorem~\ref{thm:geo}.

\subsection{Main definition}\label{ssec:noncomdef}

In order to get character sheaves that correspond to characters in $G(k)^*$, we must discard those
character sheaves whose restriction to the derived subgroup is nontrivial.
Recall from Section~\ref{sec:defs} that we refer to objects in category $\CS(G)$, defined as in \cite{cunningham-roe:13a}*{Def. 1.1}, as linear character sheaves  when $G$ is smooth but not necessarily commutative.

\begin{proposition} \label{prop:Gder_triv}
Suppose $G$ is a smooth group scheme and $(\cs{L},\mu) \in \CS(G)$.
Then the restriction of $\cs{L}$ to $G\der$ is trivial if and only if $(\cs{L},\mu) \cong q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$, for some $(\cs{L}\ab,\mu\ab) \in \CS(G\ab)$.
\end{proposition}

Proposition~\ref{prop:Gder_triv} will be proven in Section \ref{ssec:proof1}.

@@

\begin{corollary}\label{cor:Gab}
%Let $G$ be a smooth group scheme over a finite field $\Fq$. Let $\CCS(G)$ be the category of {\it commutative characters sheaves} on $G$; see Definition~\ref{def:CCS}.
Category $\CCS(G)$ is monoidal and 
\[
\CCSiso{G} \iso \Hom(G\ab(\Fq),\EEx),
\]
canonically.
Every map in $\CCS(G)$ is either trivial or an isomorphism, and the automorphism group of any object in $\CCS(G)$ is $\Hom(\pi_0(G\ab)_{\Frob{}},\EEx)$, canonically.
\end{corollary}

@@

%\begin{proof}[Proof of Proposition~\ref{prop:Gder_triv}] 
By \cite{demazure:SGA3-VIA}*{Thm. 3.2}, the abelianization $G\ab\ceq G/G\der$ exists in schemes over $\Fq$ and the quotient $q: G \to G\ab$ is a $G\der$-torsor in the fppf topology.
%
To simplify notation below, set $H=G\der$
and let $j : H\hookrightarrow G$ be the inclusion. 
With reference to \eqref{qH} and Section~\ref{ssec:closed}, consider the following diagram.
%\[
%\begin{tikzcd}
%{} & G\times G \arrow{d}{c} &&&\\
%1 \arrow{r} & H \arrow{r}{j} & G \arrow{r}{q} & G\ab \arrow{r} & 1
%\end{tikzcd}
%\]
%gives
\[
\begin{tikzcd}
\CS(H) & \arrow{l}[swap]{j^*} \CS(G) &\arrow{l}[swap]{q^*} 
\CS(G\ab) \arrow{dl}{\text{equiv}}[swap]{q_{H}^*} \\
 \CS'(G)  \arrow{r}{\beta \mapsto \nu}[swap]{\text{equiv}} &  \CS_{H}(G) \arrow{u}{\text{forget}}  & 
\end{tikzcd}
\]
%
Here we use the notation $\CS'(G)$ from Section~\ref{ssec:noncomdef} and the observation that $\CS'(G)$ is precisely the category $\CS^{H}(G)$ from Section~\ref{ssec:closed} for $H=G\der$.
%
Since the quotient $q : G \to G\ab$ is an $H$-torsor in the fppf topology \cite{demazure:SGA3-VIA}*{Thm. 3.2}, 
it is also an $H$-torsor in the fpqc topology.
It now follows from Lemma~\ref{lem:torsor} that the comparison functor $\CS(G\ab)\to \CS_{H}(G)$ is an equivalence.
%\todo{Did I compare these topologies correctly? fpqc is finer than fppf}
%
On the other hand, by Lemma~\ref{lem:beta}, $\CS'(G) \to \CS_{H}(G)$, defined by $(\cs{L},\mu,\beta)\mapsto (\cs{L},\mu, \nu)$, is an equivalence.

Now, suppose $(\cs{L},\mu)\in \CS(G)$ and there is an isomorphism $\beta : \cs{L}\vert_{H} \to (\EE)_{H}$ in $\CS(H)$.
Then $(\cs{L},\mu,\beta) \in \CS'(G)$.
We have just seen that $\CS'(G)$, $\CS_H(G)$ and $\CS(G\ab)$ are equivalent categories.
Let $(\cs{L},\mu,\beta) \mapsto (\cs{L},\mu,\nu)$ under the equivalence $\CS'(G) \to \CS_H(G)$ described above.
Then $(\cs{L},\mu,\nu) \iso q_H^*(\cs{L}\ab,\mu\ab)$ in $\CS_H(G)$, for some $(\cs{L}\ab,\mu\ab)\in \CS(G\ab)$.
Applying the forgetful functor $\CS_H(G)\to \CS(G)$ to this isomorphism, it follows that $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$, as desired.

Conversely, suppose $(\cs{L},\mu)\in \CS(G)$ and $(\cs{L},\mu) \iso q^*(\cs{L}\ab,\mu\ab)$ in $\CS(G)$.
Then $j^*(\cs{L},\mu) \iso j^*q^*(\cs{L}\ab,\mu\ab)$ in $\CS(H)$. 
Since $g\circ j = 1$, it follows that $\cs{L}\vert_{G\der} \to (\EE)_{H}$ in $\CS(H)$.
This completes the proof of Proposition~\ref{prop:Gder_triv}.
%\end{proof}
